# Descriptive Stats V

## Concepts

### Quantiles and Percentiles {.unnumbered}

A **quantile** is a location within a set of ranked numbers (or distribution), below which a certain proportion, $q$, of that set lie. Ex: 0.25 of the data lies below the 0.25 quantile.

**Percentiles** express quantiles in percentage form. Ex: 25% of the data lies below the 25th percentile. To calculate a percentile:

-   Sort the data in ascending order.

-   Compute the location of the percentile desired using $L_{p}=\frac{(n+1)P}{100}$ where $L_{p}$ is the location of the $P_{th}$ percentile, and $P$ is the percentile desired.

-   The value at $L_p$, is the the $P_{th}$ percentile.

### Chevyshev's Theorem and Empirical Rule {.unnumbered}

**Chevyshev's Theorem** states that at least $1-1/z^2$% of the data lies between $z$ standard deviations from the mean. This result does not depend on the shape of the distribution.

The **Empirical Rule** or ($68$,$95$,$99.7$ rule) states that $68$%, $95$%, and $99.7$% of the data lies between $1$, $2$, and $3$ standard deviations from the mean respectively. The rule depends on the data being normally distributed.

### Five Point Summary and Outliers {.unnumbered}

A popular way to summarize data is by calculating the minimum, first quartile, median, third quartile and maximum (five point summary).

The **interquartile range** (IQR) is the difference between the third quartile and the first quartile.

**Outliers** are extreme deviations from the mean. They are values that are not "normal". To calculate outliers:

-   Use a **z-score** to measure the distance from the mean in units of standard deviation. $z_{i}=\frac{x_i-\bar{x}}{s_x}$. $z$-scores above $3$ are suspected outliers.

-   Calculate $Q_1-1.5(IQR)$ and $Q_3+1.5(IQR)$, where $Q_1$ is the first quartile, $Q_3$ is the third quartile, and $IQR$ is the interquartile range. If $x_i$ is less than $Q_1-1.5(IQR)$ or greater than $Q_3+1.5(IQR)$, then it is considered an outlier.

A **box plot** is a graph that shows the five point summary, outliers (if any), and the distribution of data.

To determine if the data is skewed, calculate the **Pearson's Coefficient of Skew**. $Sk=\frac{3(\bar {x}- Median)}{s_x}$. The distribution is skewed to the left if $Sk<0$, skewed to the right is $Sk>0$, and symmetric if $Sk=0$.

### Useful R Functions {.unnumbered}

The `quantile()` function returns the five point summary when no arguments are specified. For a specific quantile, specify the *probs* argument.

The `boxplot()` command returns a box plot for a vector of values.

## Exercises

The following exercises will help you practice other statistical measures. In particular, the exercises work on:

-   Constructing a five point summary and a boxplot.

-   Applying Chebyshev's Theorem.

-   Identifying skewness.

-   Identifying outliers.

Answers are provided below. Try not to peak until you have a formulated your own answer and double checked your work for any mistakes.

### Exercise 1 {.unnumbered}

For the following exercises, make your calculations by hand and verify results using R functions when possible.

1.  Use the following observations to calculate the minimum, the first, second and third quartiles, and the maximum. Are there any outliers? Find the IQR to answer the question.

    |     |     |     |     |     |     |     |
    |:---:|:---:|:---:|:---:|:---:|:---:|:---:|
    |  3  | 10  |  4  |  1  |  0  | 30  |  6  |

2.  Confirm your finding of an outlier by calculating the $z$-score. Is $30$ an outlier when using a $z$-Score?

3.  Use Chebyshev's theorem to determine what percent of the data falls between the $z$-score found in $2$.

### Exercise 2 {.unnumbered}

You will need the **Stocks** data set to answer this question. You can find this data at https://jagelves.github.io/Data/Stocks.csv The data is a sample of daily stock prices for ticker symbols TSLA (Tesla), VTI (S&P 500) and GBTC (Bitcoin).

1.  Construct a boxplot for Stock A. Is the data skewed or symmetric?
2.  Create a histogram of the data. Include a vertical line for the mean and median. Explain how the mean and median indicates a skew in the data. Calculate the skewness statistic to confirm your result.
3.  Use a line chart to plot your data. Can you explain why the data has a skew?

### Exercise 3 {.unnumbered}

You will need the **mtcars** data set to answer this question. This data set is part of R. You don't need to download any files to access it.

1.  Construct a boxplot for the *hp* variable. Write a command in R that retrieves the outlier. Which car is the outlier?
2.  Create a histogram of the data. Is the data skewed? Include a vertical line for the mean and median. Calculate the skewness statistic to confirm your result.
3.  Transform the data by taking a natural logarithm. Specifically, create a new variable called *Loghp*. Repeat the procedure in 2. Is the skew still there?

## Answers

### Exercise 1 {.unnumbered}


1.  The minimum is $0$, the first quartile is $2$, second quartile is $4$, third quartile is $8$, and maximum is $30$. $30$ is an outlier since it is beyond $Q_{3}+1.5*IQR$.


Quartiles are calculated using the percentile formula $(n+1)P/100$. The data set has seven numbers. The first quartile's location is $8/4=2$, the second quartile's location is $8/2=4$ and the third quartile's location is $24/4=6$. The values at these location, when data is organized in ascending order, are $1$, $4$, and $10$.

In R we can get the five number summary by using the `quantile()` function. Since there are various rules that can be used to calculate percentiles, we specify type $6$ to match our rules.

```{r}
Ex1<-c(3,10,4,1,0,30,6)
quantile(Ex1,type = 6)
```

The interquartile range is needed to determine if there are any outliers. The $IQR$ for this data set is $Q_{3}-Q_{1}=9$. This reveals that $30$ is and outlier, since $10+1.5*9=23.5$. Everything beyond $23.5$ is an outlier.


2.  If we use the $z$-score instead we find that $30$ is not an outlier since the $z$-score is $Z_{30}=2.15$. This observation is only $2.15$ standard deviations away from the mean.


In R we can make a quick calculation of the $z$-Score to confirm our results. The $z$-score is given by $Z_{i}=\frac{x_{30}-\mu}{\sigma}$.

```{r}
(Z30<-(30-mean(Ex1))/sd(Ex1))
```


3.  Chebyshev's theorem states that $1-\frac{1}{z_{2}}$ of the data lies between $z$ standard deviation from the mean.


Substituting the $z$-score found in 2. we get $78.34$%. In R:

```{r}
1-1/(Z30)^2
```

### Exercise 2 {.unnumbered}


1.  The data is skewed to the right.


Start by loading the data set:

```{r}
StockPrices<-read.csv("https://jagelves.github.io/Data/Stocks.csv")
```

To construct the boxplot in R, use the `boxplot()` command.

```{r}
boxplot(as.numeric(StockPrices$VTI),axes=F, ylim=c(120,260),
        cex=1.5, col="#F5F5F5",pch=21,bg="red")
axis(side=1, labels=c("VTI"), at=seq(1))
axis(side=2, labels=TRUE, at=seq(140,260,20),font=1,las=1)
```

The boxplot shows that there are no outliers. The data also looks like it has a slight skew to the right.


2.  The mean is more sensitive to outliers than the median. Hence, when the data is skewed to the right we expect that the mean is larger than the median.


Let's construct a histogram in R to search for skewness.

```{r}
hist(StockPrices$VTI,main="", ylim=c(0,40), 
     xlab="Prices", col="#F5F5F5")
abline(v=mean(StockPrices$VTI),col="black",lwd=2)
abline(v=median(StockPrices$VTI),col="darkgrey",lwd=2)
legend(x = "topright",          
       legend = c("Mean", "Median"),  
       lty = c(1, 1),           
       col = c("black", "darkgrey"),         
       lwd = 2,
       bty="n")    
```

The lines are close to each other but the mean is slighlty larger than the median. Let's confirm with the skewness statistic $3(mean-median)/sd$.

```{r}
(skew<-3*(mean(StockPrices$VTI-median(StockPrices$VTI))/sd(StockPrices$VTI)))
```

This indicates that there is a slight skew to the right of the data.


3.  The line chart indicates that the data has a downward trend in the early periods. This creates a few points that are large. In later periods the stock price stabilizes to lower levels.


```{r}
plot(y=StockPrices$VTI,x=seq(1,length(StockPrices$VTI)),
     type="l", ylab="Prices", xlab="Period", axes=F)
axis(side=1, labels=TRUE, at=seq(0,250,50),font=1,las=1)
axis(side=2, labels=TRUE, at=seq(0,300,20),font=1,las=1)
```

### Exercise 3 {.unnumbered}


1.  The outlier is the Masserati Bora. The horse power is $335$.


In R we can construct a boxplot with the following command:

```{r}
boxplot(mtcars$hp,axes=F, ylim=c(0,400),
        cex=1.5, col="#F5F5F5",pch=21,bg="red")
axis(side=1, labels=c("HP"), at=seq(1))
axis(side=2, labels=TRUE, at=seq(0,400,50),font=1,las=1)
```

From the graph it seems like the outlier is beyond a horsepower of 275. Let's write an R command to retrieve the car.

```{r}
mtcars[mtcars$hp>275,]
```

It's the Masserati Bora!


2.  The histogram looks skewed to the right. This is confirmed by the estimation of a Pearson coefficient fo skewness of $1.04$.


In R we can construct a histogram with vertical lines for the mean and median wit the following code:

```{r}
hist(mtcars$hp,main="", ylim=c(0,12), xlab="Horse Power",
     col="#F5F5F5")
abline(v=mean(mtcars$hp),col="black",lwd=2)
abline(v=median(mtcars$hp),col="darkgrey",lwd=2)
legend(x = "topright",          
       legend = c("Mean", "Median"),  
       lty = c(1, 1),           
       col = c("black", "darkgrey"),         
       lwd = 2,
       bty="n")    
```

The histogram looks skewed to the right. Pearson's Coefficient of Skewness is:

```{r}
(SkewHP<-3*(mean(mtcars$hp)-median(mtcars$hp))/sd(mtcars$hp))
```


3.  The skew is still there, but the distribution now look more symmetrical and the Skew coefficient has decreased to $0.44$.


In R we can create an new variable that captures the log transformation. The `log()` function takes the natural logarithm of a number or vector.

```{r}
LogHP<-log(mtcars$hp)
```

Let's use this new variable to create our histogram:

```{r}
hist(LogHP,main="", ylim=c(0,12), xlab="Horse Power", 
     col="#F5F5F5")
abline(v=mean(LogHP),col="black",lwd=2)
abline(v=median(LogHP),col="darkgrey",lwd=2)
legend(x = "topright",          
       legend = c("Mean", "Median"),  
       lty = c(1, 1),           
       col = c("black", "darkgrey"),         
       lwd = 2,
       bty="n")    
```

The mean and the variance now look closer together. The tail of the distribution (skew) now also looks diminished. The Skewness coefficient has decreased significantly:

```{r}
(SkewLogHP<-3*(mean(LogHP)-median(LogHP))/sd(LogHP))
```
