```{=html}
<style>
details {
  background-color: #F8F9FA;
  padding: 5px;
  border: 1px solid #ddd; /* Light border */
  border-radius: 5px; /* Rounded corners */
  margin: 10px 0; /* Spacing */
}

details summary {
  cursor: pointer; /* Pointer cursor for interactivity */
}
</style>
```

# Inference II

A **confidence interval (CI)** provides a range of plausible values for a population parameter, along with a confidence level indicating that if the same procedure were repeated many times, that proportion of intervals would contain the true parameter. For example, a 95% CI means that if we repeated the sampling process 100 times, about 95 of the intervals would contain the true population parameter and 5 would not.

We can now define two important concepts. First, **the confidence level** is the proportion of times that the constructed interval would contain the true population parameter if the sampling procedure were repeated indefinitely. Second, the **significance level** is the probability of rejecting a true null hypothesis — also known as a Type I error (i.e., You conclude there is an effect, but there isn't or false positive). The significance level is captured by the greek letter $\alpha$ while and the confidence level is given by $1-\alpha$.

**Example:**

- **Confidence level (e.g., 95%) = 1 - significance level $(\alpha = 0.05)$.**
- **The remaining $(\alpha)$ represents the risk of error (Type I error in hypothesis testing context).**

CIs account for sampling variability and are wider for lower confidence levels or smaller samples. There are two competing goals when building a confidence interval:

- Accuracy — we want the interval to actually contain the true parameter
- Precision — we want the interval to be narrow enough to be informative

Choosing a very low $\alpha$ (like 0.01) prioritizes accuracy at the expense of precision, giving you a wide interval that is highly likely to include the true parameter but tells you very little about where exactly it is.

A lower significance level does not necessarily mean a better estimate — it means a safer but vaguer one. The goal is to choose an α that balances the risk of error with the need for a practically useful, precise estimate. That's why $\alpha=0.5$ has become the conventional standard in most fields — it strikes a reasonable balance between the two.

## Constructing Confidence Intervals for Means

For a population mean, assuming normality (via CLT), the CI is:

$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$ or, if $(\sigma)$ is unknown (common), use the t-distribution:

$$\bar{x} \pm t_{\alpha/2} \cdot \frac{s}{\sqrt{n}}$$

where $(\bar{x})$: Sample mean, $(z_{\alpha/2})$ or $(t_{\alpha/2})$: Critical value from standard normal or t-distribution (e.g., 1.96 for 95% CI with large n), $(s)$: Sample standard deviation and $(n)$: Sample size. The term after $(\pm)$ is the **margin of error**.



**Example:** Suppose the average life expectancy sample mean is $\bar{x} = 78.1$ years, with population standard deviation $\sigma = 4.5$, and $n = 50$. For a 90% CI $z_{0.05} \approx 1.645$:

$$\frac{\sigma}{\sqrt{n}} = \frac{4.5}{\sqrt{50}} \approx 0.637$$

$$78.1 \pm 1.645 \times 0.637 \approx 78.1 \pm 1.05$$

Lower limit (LL): 77.05\
Upper limit (UL): 79.15

We are 90% confident that the true population mean life expectancy is between 77.05 and 79.15 years.

**Example:** A random sample of $n=25$ college students reports an average weekly study time of $\bar{x} = 14.3$ hours, with a sample standard deviation of $s = 4.2$ hours. The population standard deviation is unknown. To construct a 95% confidence interval for the true mean weekly study time we start by identifying the degrees of freedom as $df=n−1=24$.

At a $\alpha = 0.05$ the critical value is $t_{0.025, \, 24} \approx 2.064$. The standard error is given by:

$$\frac{s}{\sqrt{n}} = \frac{4.2}{\sqrt{25}} = \frac{4.2}{5} = 0.84$$

Hence the confidence interval is given by: 
$$14.3 \pm 1.7314.3±2.064×0.84≈14.3±1.73$$

Lower Limit (LL): 12.57
Upper Limit (UL): 16.03

We are 95% confident that the true mean weekly study time is between 12.57 and 16.03 hours.

## Constructing Confidence Intervals for Proportions

For a population proportion $(p)$, the CI is:

$$\bar{p} \pm z_{\alpha/2} \cdot \sqrt{\frac{\bar{p}(1-\bar{p})}{n}}$$ - $\bar{p}$: Sample proportion (successes / n) - Assumes normality via CLT for proportions.

**Example:** A random sample of 100 yields 40 successes, so $\bar{p} = 0.4$. For a 90% CI $z_{0.05} \approx 1.645$:

$$SE = \sqrt{\frac{0.4 \times 0.6}{100}} = \sqrt{0.0024} \approx 0.049$$

$$0.4 \pm 1.645 \times 0.049 \approx 0.4 \pm 0.081$$

Lower Limit (LL): 0.319\
Upper Limit (UL): 0.481

We are 90% confident that the true population proportion is between 31.9% and 48.1%.

## P-Hacking
A researcher wants to estimate the average body length of an ant (in mm) for a species found in a backyard. The true population mean is unknown.

The researcher collects a sample of 30 ants and computes a 95% CI:

CI: [1.1, 2.9] mm

This interval is very wide and includes values the researcher doesn't find interesting. They were hoping to show the ants are unusually large (say, larger than 2.5 mm). The result is not significant. So instead of reporting this, they start manipulating the process:

**Hack 1: Keep re-sampling**
They go out and collect new samples repeatedly until they happen to get a sample of ants that looks bigger than average — maybe they unknowingly sampled near a food source where ants are larger.

**Hack 2: Selectively remove data**
They notice 5 ants in their sample are very small. They label them "outliers" without a real scientific reason and remove them, which pulls the mean upward.

**Hack 3: Change the confidence level**
Their 95% CI includes 2.5 mm in its range, meaning the result isn't significant. So they switch to a 90% CI, which is narrower, and now the interval is:

CI: [2.6, 3.4] mm

Now it excludes the reference value of 2.5 mm and looks significant!

**Hack 4: Measure something else**
Their body length result wasn't impressive, so they start measuring head width, leg length, antenna length — 10 different things — until one of them gives a significant result, then report only that one.


Any one of these manipulations can produce a CI that looks clean and legitimate in a published report, but was actually the result of repeated tries and selective reporting. The reader has no idea.

The Honest Version Would Say "We collected one random sample, measured body length, and found a 95% CI of [1.1, 2.9] mm. The wide interval reflects natural variability in ant size and our sample size of 30." That is far less exciting — but scientifically valid.


P-hacking with ant sizes illustrates that the problem isn't always malicious. Researchers can convince themselves their re-sampling or outlier removal is justified. But every extra "try" inflates the chance of a false result — and the published CI ends up being a lucky shot dressed up as a reliable estimate.


## Summary

Statistical inference bridges the gap between samples and populations through sampling distributions and the CLT. Confidence intervals offer a practical way to express uncertainty in estimates, essential for business decisions like market analysis or quality control. Remember, higher confidence levels widen intervals, and larger samples narrow them, improving precision.

## Useful R Functions

The `qnorm()` and `qt()` functions calculate quartiles for the normal and $t$ distributions, respectively.

The `if()` function creates a conditional statement in R.

## Exercises

The following exercises will help you test your knowledge on Statistical Inference. In particular, the exercises work on:

- Simulating confidence intervals.
- Estimating confidence intervals in R.
- Estimating confidence intervals for proportions.

Try not to peek at the answers until you have formulated your own answer and double-checked your work for any mistakes.

### Exercise 1 {.unnumbered}

In this exercise you will be simulating confidence intervals.

1.  Set the seed to $9$. Create a random sample of 1000 data points and store it in an object called *Population*. Use the exponential distribution with rate of $0.02$ to generate the data. Calculate the mean and standard deviation of *Population* and call them *PopMean* and *PopSD* respectively. What are the mean and standard deviation of *Population*?

    <details>

    <summary>Answer</summary>

    *The mean of Population is 48.61. The standard deviation is 47.94.*

    *Start by generating values from the exponential distribution. You can use the `rexp()` function in R to do this. Setting the seed to 9 yields:*

    ```{r}
    set.seed(9)
    Population <- rexp(1000, 0.02)
    ```

    *The population mean is:*

    ```{r}
    (PopMean <- mean(Population))
    ```

    *The standard deviation is:*

    ```{r}
    (PopSD <- sd(Population))
    ```

    </details>

2.  Create a for loop (with 10,000 iterations) that takes a sample of 50 points from *Population*, calculates the mean, and then stores the result in a vector called *SampleMeans*. What is the mean of the *SampleMeans*?

    <details>

    <summary>Answer</summary>

    *The mean is very close to the population mean 48.83. The standard deviation is 6.83.*

    *In R you can use a for loop to create the vector of sample means.*

    ```{r}
    nrep <- 10000
    SampleMeans <- c()
    for (i in 1:nrep){
      x <- sample(Population, 50, replace = T)
      SampleMeans <- c(SampleMeans, mean(x))
    }
    ```

    *The mean of SampleMeans is:*

    ```{r}
    (xbar <- mean(SampleMeans))
    ```

    *The standard deviation is:*

    ```{r}
    (Standard <- sd(SampleMeans))
    ```

    </details>

3.  Create a $90$% confidence interval using the first data point in the *SampleMeans* vector. Does the confidence interval include *PopMean*?

    <details>

    <summary>Answer</summary>

    *The confidence interval is \[47.71, 70.17\]. Since the population mean is equal to 48.61, the confidence interval does include the population mean.*

    *Let's construct the upper and lower limits of the interval in R.*

    ```{r}
    (ll <- SampleMeans[1] + qnorm(0.05) * Standard)
    (ul <- SampleMeans[1] - qnorm(0.05) * Standard)
    ```

    </details>

4.  Now take the minimum of the *SampleMeans* vector. Create a new $90$% confidence interval. Does the interval include *PopMean*? Out of the $10,000$ intervals that you could construct with the vector *SampleMeans*, how many would you expect to include *PopMean*?

    <details>

    <summary>Answer</summary>

    *The confidence interval is \[14.86, 37.32\]. This interval does not include the population mean of 48.61. Out of the 10,000 confidence intervals, one would expect about 9,000 to include the population mean.*

    *Let's find the confidence interval limits using R.*

    ```{r}
    (Minll <- min(SampleMeans) + qnorm(0.05) * Standard)
    (Minul <- min(SampleMeans) - qnorm(0.05) * Standard)
    ```

    *We can confirm in R that about 9,000 of the intervals include PopMean. Once more, let's use a for loop to construct confidence intervals for each element in SampleMeans and check whether the PopMean is included. The count variable keeps track of how many intervals include the population mean.*

    ```{r}
    count = 0

    for (i in SampleMeans){
      (ll <- i + qnorm(0.05) * Standard)
      (ul <- i - qnorm(0.05) * Standard)
      if (PopMean <= ul & PopMean >= ll){
        count = count + 1
      }
    }

    count
    ```

    </details>

### Exercise 2 {.unnumbered}

1.  A random sample of $24$ observations is used to estimate the population mean. The sample mean is $104.6$ and the standard deviation is $28.8$. The population is normally distributed. Construct a $90$% and $95$% confidence interval for the population mean. How does the confidence level affect the size of the interval?

    <details>

    <summary>Answer</summary>

    *The 90% confidence interval is \[94.52, 114.67\] and the 95% confidence interval is \[92.68, 116.76\]. The larger the confidence level, the larger the interval.*

    *Let's construct the intervals using R. Since the population standard deviation is unknown we will use the t-distribution. The interval is constructed as (*\bar{x} \pm t\_{\alpha/2} \frac{s}{\sqrt{n}}).

    ```{r}
    (ul90 <- 104.6 - qt(0.05, 23) * 28.8 / sqrt(24))
    (ll90 <- 104.6 + qt(0.05, 23) * 28.8 / sqrt(24))
    ```

    *For the 95% confidence interval we adjust the significance level accordingly.*

    ```{r}
    (ul95 <- 104.6 - qt(0.025, 23) * 28.8 / sqrt(24))
    (ll95 <- 104.6 + qt(0.025, 23) * 28.8 / sqrt(24))
    ```

    </details>

2.  A random sample from a normally distributed population yields a mean of $48.68$ and a standard deviation of $33.64$. Compute a $95$% confidence interval assuming a) that the sample size is $16$ and b) the sample size is $25$. What happens to the confidence interval as the sample size increases?

    <details>

    <summary>Answer</summary>

    *The confidence interval for a sample size of 16 is \[30.75, 66.61\]. The confidence interval when the sample size is 25 is \[34.79, 62.57\]. As the sample size gets larger, the confidence interval gets narrower and more precise.*

    *Let's use R again to calculate the confidence interval. For a sample size of 16 the interval is:*

    ```{r}
    (ul16 <- 48.68 - qt(0.025, 15) * 33.64 / sqrt(16))
    (ll16 <- 48.68 + qt(0.025, 15) * 33.64 / sqrt(16))
    ```

    *Increasing the sample size to 25 yields:*

    ```{r}
    (ul25 <- 48.68 - qt(0.025, 24) * 33.64 / sqrt(25))
    (ll25 <- 48.68 + qt(0.025, 24) * 33.64 / sqrt(25))
    ```

    </details>

### Exercise 3 {.unnumbered}

You will need the **sleep** data set for this problem. The data is built into R, and displays the effect of two sleep inducing drugs on students. Calculate a $95$% confidence interval for group 1 and for group 2. Which drug would you expect to be more effective at increasing sleeping times?

<details>

<summary>Answer</summary>

*The 95% confidence interval for group 1 is \[-0.53, 2.03\]. Let's first calculate the standard error for group 1.*

```{r}
(se1 <- sd(sleep$extra[sleep$group == 1]) / sqrt(length(sleep$extra[sleep$group == 1])))
```

*We can now use the standard error to estimate the lower and upper limits of the confidence interval.*

```{r}
(ll1 <- mean(sleep$extra[sleep$group == 1]) + qt(0.025, 9) * se1)
(ul1 <- mean(sleep$extra[sleep$group == 1]) - qt(0.025, 9) * se1)
```

*The 95% confidence interval for group 2 is \[0.90, 3.76\].Let's repeat the procedure for group 2. Start by finding the standard error.*

```{r}
(se2 <- sd(sleep$extra[sleep$group == 2]) / sqrt(length(sleep$extra[sleep$group == 2])))
```

*Using the standard error we can complete the confidence interval.*

```{r}
(ll2 <- mean(sleep$extra[sleep$group == 2]) + qt(0.025, 9) * se2)
(ul2 <- mean(sleep$extra[sleep$group == 2]) - qt(0.025, 9) * se2)
```

*Drug 2 is more effective. Drug 2 does not include zero in the interval, and the interval is to the right of zero. It is unlikely that drug 2 has no effect on students' sleeping time. Additionally, Drug 2's mean increase in sleeping hours is 2.33 vs. 0.75 for drug 1.*

</details>

### Exercise 4 {.unnumbered}

1.  A random sample of $100$ observations results in $40$ successes. Construct a $90$% and $95$% confidence interval for the population proportion. Can we conclude at either confidence level that the population proportion differs from $0.5$?

    <details>

    <summary>Answer</summary>

    *The 90% and 95% confidence intervals are \[0.319, 0.481\], and \[0.304, 0.496\] respectively. Since they do not include 0.5, we can conclude that the population proportion is significantly different from 0.5.*

    *We can create an object that stores the sample proportion and sample in R:*

    ```{r}
    (p <- 0.4)
    (n <- 100)
    ```

    *The 90% confidence interval is given by:*

    ```{r}
    (Ex1ll90 <- p + qnorm(0.05) * sqrt(p * (1 - p) / 100))
    (Ex1ul90 <- p - qnorm(0.05) * sqrt(p * (1 - p) / 100))
    ```

    *The 95% confidence interval is:*

    ```{r}
    (Ex1ll95 <- p + qnorm(0.025) * sqrt(p * (1 - p) / 100))
    (Ex1ul95 <- p - qnorm(0.025) * sqrt(p * (1 - p) / 100))
    ```

    </details>

2.  You will need the **HairEyeColor** data set for this problem. The data is built into R, and displays the distribution of hair and eye color for $592$ statistics students. Construct a $95$ confidence interval for the proportion of Hazel eye color students.

    <details>

    <summary>Answer</summary>

    *The 95% confidence interval is \[0.128, 0.186\].*

    *The data can easily be viewed by calling `HairEyeColor` in R.*

    ```{r}
    HairEyeColor
    ```

    *Note that there are three dimensions to this table (Hair, Eye, Sex). We can calculate the proportion of Hazel eye colored students with the following command that makes use of indexing:*

    ```{r}
    (p <- (sum(HairEyeColor[, 3, 1]) + sum(HairEyeColor[, 3, 2])) / sum(HairEyeColor))
    ```

    *Now we can use this proportion to construct the intervals. Recall that for proportions the interval is calculated by (*\bar{p} \pm z\_{\alpha/2} \sqrt{\frac{\bar{p}(1 - \bar{p})}{n}}). The 95% confidence interval is given by:

    ```{r}
    (Ex2ll95 <- p + qnorm(0.025) * sqrt(p * (1 - p) / 592))
    (Ex2ul95 <- p - qnorm(0.025) * sqrt(p * (1 - p) / 592))
    ```

    </details>

### Exercise 5 {.unnumbered}

The **2024 Formula 1 Crypto.com Miami Grand Prix** was held on May 5, 2024 at the
Miami International Autodrome. Lando Norris won the race for McLaren in his first-ever
F1 victory. The table below shows the **fastest lap time** recorded by each of the 20
drivers during the race.

| Driver | Team | Fastest Lap |
|---|---|---|
| Oscar Piastri | McLaren Mercedes | 1:30.634 |
| Alexander Albon | Williams Mercedes | 1:30.849 |
| Sergio Perez | Red Bull Racing | 1:30.855 |
| Carlos Sainz | Ferrari | 1:30.928 |
| Lando Norris | McLaren Mercedes | 1:30.980 |
| Charles Leclerc | Ferrari | 1:31.084 |
| Lewis Hamilton | Mercedes | 1:31.233 |
| Max Verstappen | Red Bull Racing | 1:31.261 |
| Lance Stroll | Aston Martin | 1:31.588 |
| Yuki Tsunoda | RB Honda RBPT | 1:31.682 |
| Fernando Alonso | Aston Martin | 1:31.727 |
| Kevin Magnussen | Haas Ferrari | 1:31.774 |
| George Russell | Mercedes | 1:31.921 |
| Nico Hulkenberg | Haas Ferrari | 1:31.941 |
| Zhou Guanyu | Kick Sauber | 1:31.991 |
| Esteban Ocon | Alpine Renault | 1:32.037 |
| Pierre Gasly | Alpine Renault | 1:32.055 |
| Valtteri Bottas | Kick Sauber | 1:32.098 |
| Daniel Ricciardo | RB Honda RBPT | 1:32.122 |
| Logan Sargeant | Williams Mercedes | 1:33.452 |

*Source: Formula 1 Official Results, formula1.com*

---

We can store the data into a tibble and convert the lap times
to seconds with the code below:

```{r}
library(tidyverse)

   miami_2024 <- tibble(
     position    = 1:20,
     driver      = c("Oscar Piastri", "Alexander Albon", "Sergio Perez",
                     "Carlos Sainz", "Lando Norris", "Charles Leclerc",
                     "Lewis Hamilton", "Max Verstappen", "Lance Stroll",
                     "Yuki Tsunoda", "Fernando Alonso", "Kevin Magnussen",
                     "George Russell", "Nico Hulkenberg", "Zhou Guanyu",
                     "Esteban Ocon", "Pierre Gasly", "Valtteri Bottas",
                     "Daniel Ricciardo", "Logan Sargeant"),
     team        = c("McLaren Mercedes", "Williams Mercedes", "Red Bull Racing",
                     "Ferrari", "McLaren Mercedes", "Ferrari",
                     "Mercedes", "Red Bull Racing", "Aston Martin",
                     "RB Honda RBPT", "Aston Martin", "Haas Ferrari",
                     "Mercedes", "Haas Ferrari", "Kick Sauber",
                     "Alpine Renault", "Alpine Renault", "Kick Sauber",
                     "RB Honda RBPT", "Williams Mercedes"),
     fastest_lap = c("1:30.634", "1:30.849", "1:30.855", "1:30.928", "1:30.980",
                     "1:31.084", "1:31.233", "1:31.261", "1:31.588", "1:31.682",
                     "1:31.727", "1:31.774", "1:31.921", "1:31.941", "1:31.991",
                     "1:32.037", "1:32.055", "1:32.098", "1:32.122", "1:33.452"),
     lap_seconds = c(90.634, 90.849, 90.855, 90.928, 90.980,
                     91.084, 91.233, 91.261, 91.588, 91.682,
                     91.727, 91.774, 91.921, 91.941, 91.991,
                     92.037, 92.055, 92.098, 92.122, 93.452)
   )

```
1. What is the sample mean and sample standard deviation?

   <details>
   <summary>Answer</summary>

   *The sample mean is 91.611 seconds (1:31.611) and the sample standard deviation is
   0.656 seconds. Since the true population standard deviation is unknown, we will use
   the t-distribution to construct confidence intervals.*

The sample size is:
```{r}
   (n <- nrow(miami_2024))
```

The mean:
```{r}
   (xbar <- mean(miami_2024$lap_seconds))
```

and the Standard Deviation
```{r}
   (s <- sd(miami_2024$lap_seconds))
```
   </details>

2. Construct a **90%** and a **95%** confidence interval for the true mean fastest lap
   time at the Miami circuit. Use the `t.test()` function to verify your results.

   <details>
   <summary>Answer</summary>

   *The 90% confidence interval is [1:31.357, 1:31.864] and the 95% confidence interval
   is [1:31.304, 1:31.918]. Since $n = 20$, we use the t-distribution with $df = n - 1 = 19$ degrees of freedom.*

   *The standard error is:*

```{r}
   (se <- s / sqrt(n))
```

   *For the 90% CI ($\alpha = 0.10$), $t_{0.05, \, 19} \approx 1.729$:*

The lower bound is:
```{r}
   (ll90 <- xbar + qt(0.05, 19) * se)
```

and the upper:
```{r}
   (ul90 <- xbar - qt(0.05, 19) * se)
```

   *For the 95% CI ($\alpha = 0.05$), $t_{0.025, \, 19} \approx 2.093$:*

The lower:
```{r}
   (ll95 <- xbar + qt(0.025, 19) * se)
```

   
The upper:
```{r}
   (ul95 <- xbar - qt(0.025, 19) * se)
```

   *We can verify using `t.test()`:*

```{r}
t.test(miami_2024$lap_seconds, conf.level = 0.95)
```

   </details>

---

3. Logan Sargeant's fastest lap of 1:33.452 (93.452 sec) is noticeably slower than
   the rest of the field. Remove his lap time from the data and recalculate the 95%
   confidence interval. How does removing this outlier affect the interval, and what
   does this illustrate about the sensitivity of confidence intervals to extreme values?

   <details>
   <summary>Answer</summary>

   *Without Sargeant's lap time, the 95% CI narrows to approximately [91.189, 91.630],
   with a new mean of 91.409 seconds. The interval shifts downward and becomes
   substantially narrower — the margin of error drops because the sample standard
   deviation falls sharply once the outlier is removed. This illustrates that confidence
   intervals are sensitive to extreme observations, particularly in small samples. This
   is also a reminder of why removing data points should only be done with a clear
   justification — in this case, Sargeant set his fastest lap on lap 15 during an
   early pit phase, making it a genuinely unrepresentative time.*

```{r}
miami_no_sar <- miami_2024 |> filter(driver != "Logan Sargeant")

   t.test(miami_no_sar$lap_seconds, conf.level = 0.95)
```

   </details>